---
title: "Introduction to BacArena"
author: "Eugen Bauer and Johannes Zimmermann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


BacArena is a Systems Biology project to simulate and analyse the astonishing behavior of microbial communities. Just take one or some of the widespread metabolic network reconstructions^[http://gcrg.ucsd.edu/InSilicoOrganisms/OtherOrganisms], put them into the *arena*, and follow up their interactions!
Some features
- Growth model 
- Substrate diffusion
- Movement
- Simulation of internal metabolism

## Installing

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Setting up the arena

```{r, fig.show='hold'}
load("data/Bcoli_model.R")
ecore = model
bace1 = Bac(model=ecore, deathrate=0.1, duplirate=1.5, growthlimit=0.05, growtype="exponential",
           speed=2, type="ecore1", lyse=T)
arena = Arena(n=100, m=100, tstep=1)
addOrg(arena, bace1, amount=10)
addSubs(arena, smax=2000)

```


## First simulation

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
evalsim <- simulate(arena, time=20)
plotCurves(evalsim, remove=F, retdata = F)
```

